---
title: "ePRINT peaks analysis"
author: "`r Sys.getenv('USER')`"
date: "`r Sys.Date()`"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Configuration

All experiment-specific paths and labels live here. Adapt before knitting.
```{r config}
library(Biostrings)
library(GenomicRanges)
library(BSgenome.Hsapiens.UCSC.hg38)
library(rtracklayer)
library(pheatmap)
library(universalmotif)
library(ape)
library(ChIPseeker)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)

txdb <- TxDb.Hsapiens.UCSC.hg38.knownGene

# Root of the peak-processing output directory
peak_dir <- "/path/to/your/experiment/peak_processing/"

# Directory for figure/table outputs (used when writing results for publication)
fig_dir  <- "/path/to/your/figures/output/"

# Path to the pre-built combined PWM RData object
pwm_rdata <- "/path/to/your/resources/motifs/all_combined_pwms.RData"

# Path to POSTAR3 CLIP ranges RData object (for cross-validation)
clip_rdata <- "/path/to/your/resources/clip_data/postar3/all_clip_ranges.RData"

# Path to splicing landscape RData objects
splicing_background_rdata <- "/path/to/your/splicing/background_events.RData"
splicing_landscape_rdata  <- "/path/to/your/splicing/all_landscapes.RData"

# Condition labels — must match those used in peaks_r_treat.R / peak_processing.sh
cond1 <- "condition1"   # e.g. "noDOX"
cond2 <- "condition2"   # e.g. "DOX"

# Number of bedtools-shuffled background sets generated per condition (Step 5 of
# peak_processing.sh). Shuffled files are named <cond>_shuffled1.bed … <cond>_shuffledN.bed
N_SHUFFLES <- 5

# Extension (bp) applied to peaks when computing overlaps and extracting sequences
PEAK_EXT <- 50

# Fold-change threshold for calling condition-exclusive peaks
# (peak is exclusive if its mean_lfc is > FC_EXCLUSIVE_THRESH * the other condition's)
FC_EXCLUSIVE_THRESH <- 2

# Replacement for Inf in -log10(p) columns
INF_SENTINEL <- 400

load(pwm_rdata)
```

---

## 1  Loading reproducible merged peaks
```{r load-peaks}
#' Read a reprod.merged BED file and return a GRanges with score metadata
load_reprod_merged <- function(path) {
  raw <- read.table(path)
  GRanges(
    seqnames  = raw$V1,
    ranges    = IRanges(start = raw$V2, end = raw$V3),
    strand    = raw$V6,
    # Columns 7 and 9 carry log2FC from each replicate; 4,5 carry peak p-values;
    # 8 and 10 carry enrichment p-values — see peaks_r_treat.R for column layout
    mean_lfc         = rowMeans(cbind(raw$V7, raw$V9)),
    mean_peakPval    = rowMeans(cbind(raw$V4, raw$V5)),
    mean_enrichPval  = rowMeans(cbind(raw$V8, raw$V10))
  )
}

ranges_c1 <- load_reprod_merged(paste0(peak_dir, cond1, ".reprod.merged.bed"))
ranges_c2 <- load_reprod_merged(paste0(peak_dir, cond2, ".reprod.merged.bed"))

ranges_c1
ranges_c2
```

---

## 2  Peak width distributions
```{r peak-widths}
hist(width(ranges_c1), breaks = 30, main = paste(cond1, "peak widths"), xlab = "Width (bp)")
hist(width(ranges_c2), breaks = 30, main = paste(cond2, "peak widths"), xlab = "Width (bp)")

write.table(width(ranges_c1),
  file = paste0(fig_dir, "width_", cond1, ".txt"),
  quote = FALSE, row.names = FALSE, col.names = FALSE)

write.table(width(ranges_c2),
  file = paste0(fig_dir, "width_", cond2, ".txt"),
  quote = FALSE, row.names = FALSE, col.names = FALSE)
```

---

## 3  Genomic annotations
```{r annotations}
annot_priority <- c("5UTR", "3UTR", "Exon", "Intron", "Promoter", "Downstream", "Intergenic")

annot_c1 <- annotatePeak(ranges_c1, TxDb = txdb, genomicAnnotationPriority = annot_priority)
plotAnnoPie(annot_c1)
annot_c1

annot_c2 <- annotatePeak(ranges_c2, TxDb = txdb, genomicAnnotationPriority = annot_priority)
plotAnnoPie(annot_c2)
annot_c2
```

---

## 4  Cross-validation with POSTAR3 CLIP data
```{r clip-overlap}
load(clip_rdata)

#' Fraction of peaks in `query` overlapping at least one CLIP range
frac_clip_overlap <- function(query, clip_ranges, ext = PEAK_EXT) {
  o  <- findOverlaps(query + ext, clip_ranges)
  tt <- table(queryHits(o))
  v  <- integer(length(query))
  v[as.integer(names(tt))] <- as.integer(tt)
  mean(v > 0)
}

frac_c1     <- frac_clip_overlap(ranges_c1, all_clip_ranges)
frac_c2     <- frac_clip_overlap(ranges_c2, all_clip_ranges)

# Random background: pooled shuffled peaks generated in peak_processing.sh
random_peaks  <- read.delim(paste0(peak_dir, "random_peaks.bed"), header = FALSE)
random_ranges <- GRanges(
  seqnames = random_peaks$V1,
  ranges   = IRanges(start = random_peaks$V2, end = random_peaks$V3),
  strand   = random_peaks$V6
)
frac_random <- frac_clip_overlap(random_ranges, all_clip_ranges)

message(cond1,  " peaks overlapping CLIP: ", round(frac_c1     * 100, 1), "%")
message(cond2,  " peaks overlapping CLIP: ", round(frac_c2     * 100, 1), "%")
message("Random peaks overlapping CLIP:  ", round(frac_random  * 100, 1), "%")

# Permutation test: how often does a same-size random sample match or exceed
# the observed overlap fraction?
perm_test <- function(background_v, observed_frac, n_peaks, n_rep = 10000) {
  rep_fracs <- replicate(n_rep, mean(sample(background_v, size = n_peaks) > 0))
  mean(rep_fracs >= observed_frac)
}

v_random <- {
  o  <- findOverlaps(random_ranges + PEAK_EXT, all_clip_ranges)
  tt <- table(queryHits(o))
  v  <- integer(length(random_ranges))
  v[as.integer(names(tt))] <- as.integer(tt)
  v
}

message(cond1, " permutation p-value: ",
  perm_test(v_random, frac_c1, length(ranges_c1)))
message(cond2, " permutation p-value: ",
  perm_test(v_random, frac_c2, length(ranges_c2)))
```

---

## 5  Sequence extraction
```{r extract-sequences}
#' Extract peak sequences (strict and extended), write FASTA, return DNAStringSet
extract_and_write_seqs <- function(gr, label, ext = PEAK_EXT, out_dir = peak_dir) {

  seqs_strict <- getSeq(BSgenome.Hsapiens.UCSC.hg38, gr)
  names(seqs_strict) <- paste0("peak_", seq_along(seqs_strict))
  writeXStringSet(seqs_strict, filepath = paste0(out_dir, label, "_seqs.fasta"))

  seqs_ext <- getSeq(BSgenome.Hsapiens.UCSC.hg38, gr + ext)
  names(seqs_ext) <- paste0("peak_", seq_along(seqs_ext))
  writeXStringSet(seqs_ext,    filepath = paste0(out_dir, label, "_seqs_ext", ext, ".fasta"))

  invisible(list(strict = seqs_strict, extended = seqs_ext))
}

extract_and_write_seqs(ranges_c1, cond1)
extract_and_write_seqs(ranges_c2, cond2)

# Shuffled background sequences (one FASTA per shuffle set, named <cond>_shuffled<i>)
for (cond in c(cond1, cond2)) {
  for (i in seq_len(N_SHUFFLES)) {
    shuf_raw <- read.table(paste0(peak_dir, cond, "_shuffled", i, ".bed"))
    shuf_gr  <- GRanges(
      seqnames = shuf_raw$V1,
      ranges   = IRanges(start = shuf_raw$V2, end = shuf_raw$V3),
      strand   = shuf_raw$V6
    )
    extract_and_write_seqs(shuf_gr, paste0(cond, "shuf", i))
  }
}
```

---

## 6  Differential binding — condition-exclusive and common peaks
```{r differential-binding}
#' Classify peaks in `query_ranges` relative to `subject_ranges`.
#' Returns a list: $exclusive (GRanges), $common (GRanges)
classify_peaks <- function(query_ranges, subject_ranges,
                           fc_thresh = FC_EXCLUSIVE_THRESH) {

  olap     <- findOverlaps(query_ranges, subject_ranges)
  hit_idx  <- unique(queryHits(olap))

  # For each query peak that overlaps a subject peak, compare mean_lfc
  classification <- vapply(hit_idx, function(i) {
    pos     <- which(queryHits(olap) == i)
    q_score <- max(query_ranges[queryHits(olap[pos])]$mean_lfc)
    s_score <- max(subject_ranges[subjectHits(olap[pos])]$mean_lfc)
    if      (q_score > fc_thresh * s_score) "Increased"
    else if (s_score > fc_thresh * q_score) "Decreased"
    else                                     "Common"
  }, character(1))

  # Exclusive = no overlap at all + overlapping but strongly increased
  no_overlap_idx <- setdiff(seq_along(query_ranges), hit_idx)
  increased_idx  <- hit_idx[classification == "Increased"]
  exclusive_idx  <- sort(c(no_overlap_idx, increased_idx))

  common_idx <- hit_idx[classification == "Common"]

  list(
    exclusive = query_ranges[exclusive_idx],
    common    = query_ranges[common_idx]
  )
}

res_c1 <- classify_peaks(ranges_c1, ranges_c2)
res_c2 <- classify_peaks(ranges_c2, ranges_c1)

exclusive_c1 <- res_c1$exclusive
exclusive_c2 <- res_c2$exclusive
common_c1    <- res_c1$common
common_c2    <- res_c2$common

# Combine common peaks from both conditions into a single unprocessed BED
# (will be sorted and merged by bedtools in the shell)
all_common_df <- rbind(
  cbind(
    as.character(seqnames(common_c1)),
    start(common_c1), end(common_c1),
    common_c1$mean_enrichPval, common_c1$mean_lfc,
    as.character(strand(common_c1))
  ),
  cbind(
    as.character(seqnames(common_c2)),
    start(common_c2), end(common_c2),
    common_c2$mean_enrichPval, common_c2$mean_lfc,
    as.character(strand(common_c2))
  )
)

write.table(
  all_common_df,
  file      = paste0(peak_dir, "common_peaks.unprocessed.bed"),
  sep       = "\t", quote = FALSE,
  row.names = FALSE, col.names = FALSE
)

message("Written: common_peaks.unprocessed.bed")
message("  → Run bedtools sort + merge in the shell before the next chunk.")
```

---

## 7  Post-merge: load common peaks and extract sequences

Run the bedtools sort/merge step in the shell first, then continue here.
```{r common-peaks-post-merge}
common_peaks_raw <- read.table(paste0(peak_dir, "common_peaks.merged.bed"))
common_ranges    <- GRanges(
  seqnames = common_peaks_raw$V1,
  ranges   = IRanges(start = common_peaks_raw$V2, end = common_peaks_raw$V3),
  strand   = common_peaks_raw$V6
)

extract_and_write_seqs(common_ranges,  "common_ranges")
extract_and_write_seqs(exclusive_c1,   paste0(cond1, "_exclusive"))
extract_and_write_seqs(exclusive_c2,   paste0(cond2, "_exclusive"))

# Write exclusive peak BED files for downstream use
export(exclusive_c1, paste0(peak_dir, cond1, "_exclusive.bed"), "bed")
export(exclusive_c2, paste0(peak_dir, cond2, "_exclusive.bed"), "bed")
```

---

## 8  Annotations of differential peak sets
```{r annotations-differential}
annot_excl_c1   <- annotatePeak(exclusive_c1,  TxDb = txdb, genomicAnnotationPriority = annot_priority)
annot_excl_c2   <- annotatePeak(exclusive_c2,  TxDb = txdb, genomicAnnotationPriority = annot_priority)
annot_common    <- annotatePeak(common_ranges,  TxDb = txdb, genomicAnnotationPriority = annot_priority)

plotAnnoPie(annot_excl_c1);  annot_excl_c1
plotAnnoPie(annot_excl_c2);  annot_excl_c2
plotAnnoPie(annot_common);   annot_common
```

---

## 9  Motif scoring (PWM z-score heatmaps)

Run `count_pwm.R` on the cluster first to produce `pwm_colmeans_ext50.RData`.
```{r motif-heatmaps}
load(paste0(peak_dir, "pwm_colmeans_ext50.RData"))

#' Scale a PWM means matrix and remove rows with NAs
safe_scale <- function(mat) {
  s <- t(scale(mat))
  na_rows <- unique(which(is.na(s), arr.ind = TRUE)[, 1])
  if (length(na_rows) > 0) s <- s[-na_rows, ]
  s
}

s_c1      <- safe_scale(get(paste0("pwm_means_", cond1, "ext50")))
s_c1_shuf <- safe_scale(get(paste0("pwm_means_", cond1, "ext50_shuff")))
s_c1_rand <- safe_scale(get(paste0("pwm_means_", cond1, "ext50_rand")))

s_c2      <- safe_scale(get(paste0("pwm_means_", cond2, "ext50")))
s_c2_shuf <- safe_scale(get(paste0("pwm_means_", cond2, "ext50_shuff")))
s_c2_rand <- safe_scale(get(paste0("pwm_means_", cond2, "ext50_rand")))

# Restrict to motifs present in all three backgrounds
common_motifs <- Reduce(intersect, list(
  rownames(s_c1), rownames(s_c1_shuf), rownames(s_c1_rand)
))
s_c1      <- s_c1     [rownames(s_c1)      %in% common_motifs, ]
s_c1_shuf <- s_c1_shuf[rownames(s_c1_shuf) %in% common_motifs, ]
s_c1_rand <- s_c1_rand[rownames(s_c1_rand) %in% common_motifs, ]

# Hierarchical clustering on condition 1 peaks to define motif order
hc_c1 <- hclust(dist(s_c1))

# Write ordered z-score tables and the motif tree
write.tree(
  as.phylo(hc_c1),
  file = paste0(fig_dir, "tree_motifs.dnd")
)
write.table(s_c1     [hc_c1$order, ], paste0(fig_dir, "zsco_motifs_", cond1,        ".txt"), quote = FALSE, row.names = TRUE, col.names = FALSE)
write.table(s_c1_shuf[hc_c1$order, ], paste0(fig_dir, "zsco_motifs_", cond1, "_shuff.txt"), quote = FALSE, row.names = TRUE, col.names = FALSE)
write.table(s_c1_rand[hc_c1$order, ], paste0(fig_dir, "zsco_motifs_", cond1, "_rand.txt"),  quote = FALSE, row.names = TRUE, col.names = FALSE)

# Helper to draw a diverging heatmap with a zero-centred colour scale
plot_motif_heatmap <- function(mat, title = "") {
  my_colors <- colorRampPalette(c("blue", "white", "red"))(100)
  z_lim  <- max(abs(mat), na.rm = TRUE)
  breaks <- c(seq(-z_lim, 0, length.out = 51),
               seq(0,  z_lim, length.out = 51)[-1])
  pheatmap(mat, main = title, cluster_cols = FALSE,
           color = my_colors, breaks = breaks)
}

plot_motif_heatmap(s_c1,      paste(cond1, "peaks"))
plot_motif_heatmap(s_c1_shuf, paste(cond1, "shuffled"))
plot_motif_heatmap(s_c1_rand, paste(cond1, "random"))

plot_motif_heatmap(s_c2,      paste(cond2, "peaks"))
plot_motif_heatmap(s_c2_shuf, paste(cond2, "shuffled"))
plot_motif_heatmap(s_c2_rand, paste(cond2, "random"))
```

---

## 10  Overlap with the EwS splicing landscape
```{r splicing-overlap}
load(splicing_background_rdata)
load(splicing_landscape_rdata)

# Build GRanges spanning the full alternative-splicing event window for each
# event type; stacks significant events and background events into single objects

build_as_ranges <- function(se, ri, mxe, a5ss, a3ss) {
  chr  <- c(se$chr, ri$chr, mxe$chr, a5ss$chr, a3ss$chr)
  str  <- c(se$strand, ri$strand, mxe$strand, a5ss$strand, a3ss$strand)
  # A5SS / A3SS: longer exon boundaries differ by strand
  s1 <- c(
    se$upstreamES,  ri$upstreamES,  mxe$upstreamES,
    ifelse(a5ss$strand == "+", a5ss$longExonStart_0base, a5ss$flankingEE),
    ifelse(a3ss$strand == "+", a3ss$flankingEE,          a3ss$longExonStart_0base)
  )
  e1 <- c(
    se$downstreamEE,  ri$downstreamEE,  mxe$downstreamEE,
    ifelse(a5ss$strand == "+", a5ss$flankingEE,          a5ss$longExonStart_0base),
    ifelse(a3ss$strand == "+", a3ss$longExonStart_0base, a3ss$flankingEE)
  )
  as_type <- c(
    rep("se",   nrow(se)),  rep("ri",   nrow(ri)),
    rep("mxe",  nrow(mxe)), rep("a5ss", nrow(a5ss)),
    rep("a3ss", nrow(a3ss))
  )
  GRanges(seqnames = chr, ranges = IRanges(start = s1, end = e1),
          strand = str, as_type = as_type)
}

ranges_sig  <- build_as_ranges(land_se_dir, land_ri_dir, land_mxe_dir,
                                land_a5ss_dir, land_a3ss_dir)
ranges_back <- build_as_ranges(back_se,      back_ri,    back_mxe,
                                back_a5ss,    back_a3ss)

#' Fraction of AS events overlapping a peak set
frac_as_overlap <- function(as_ranges, peak_ranges, ext = PEAK_EXT) {
  o  <- findOverlaps(as_ranges, peak_ranges + ext)
  tt <- table(queryHits(o))
  v  <- integer(length(as_ranges))
  v[as.integer(names(tt))] <- as.integer(tt)
  mean(v > 0)
}

# Report overlap fractions for significant and background events
for (label in c(cond1, cond2)) {
  peaks <- switch(label, !!cond1 := exclusive_c1, !!cond2 := exclusive_c2)
  message(label, " exclusive — sig overlap:  ", round(frac_as_overlap(unique(ranges_sig),  peaks) * 100, 1), "%")
  message(label, " exclusive — back overlap: ", round(frac_as_overlap(unique(ranges_back), peaks) * 100, 1), "%")
}
message("Common peaks — sig overlap:  ", round(frac_as_overlap(unique(ranges_sig),  common_ranges) * 100, 1), "%")
message("Common peaks — back overlap: ", round(frac_as_overlap(unique(ranges_back), common_ranges) * 100, 1), "%")

# Skipped-exon subset separately (most biologically interpretable event type)
ranges_se_sig  <- GRanges(seqnames = land_se_dir$chr,
  ranges = IRanges(start = land_se_dir$upstreamES, end = land_se_dir$downstreamEE),
  strand = land_se_dir$strand)
ranges_se_back <- GRanges(seqnames = back_se$chr,
  ranges = IRanges(start = back_se$upstreamES, end = back_se$downstreamEE),
  strand = back_se$strand)

message("SE events — ", cond1, " excl overlap (sig/back): ",
  round(frac_as_overlap(ranges_se_sig,  exclusive_c1) * 100, 1), "% / ",
  round(frac_as_overlap(ranges_se_back, exclusive_c1) * 100, 1), "%")
message("SE events — ", cond2, " excl overlap (sig/back): ",
  round(frac_as_overlap(ranges_se_sig,  exclusive_c2) * 100, 1), "% / ",
  round(frac_as_overlap(ranges_se_back, exclusive_c2) * 100, 1), "%")
```



